Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-23 main.Rmd
reorganised and changed language  of abstract

abstract: |
 Recent advances in technology greatly facilitates recording and storing data at much finer temporal scales than was previously possible. As the frequency of time-oriented data increases, the number of questions about the observed variable that can be addressed by visual representation also increases. We propose some new tools to explore this type of data, which deconstruct time in many different ways. There are several classes of time deconstructions including linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Cyclic time granularities can be circular such as hour of the day, and day of the week, quasi-circular such as day of the month, and aperiodic such as public or school holidays.

  The hierarchical structure of linear granularities creates a natural nested ordering resulting in single-order-up and multiple-order-up granularities. For example, hour of the week and second of the hour are both multiple-order-up, while hour of the day and second of the minute are single-order-up.

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within a tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk]. The ideas presented in this article have been implemented in the open source R package gravitas. R code and data are available in the supplementary materials.

bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

>>>>

abstract: |
 Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes several classes of time deconstructions using linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months with respect to a baseline. Cyclic time granularities can be circular such as hour of the day, quasi-circular such as day of the month, and aperiodic such as public  holidays. The hierarchical structure of granularities creates a nested ordering. Hour of the day and second of the minute are single-order-up. Hour of the week is multiple-order-up, because it passes over day  of the week. Methods are provided for creating all possible granularities for a time index.  A recommendation algorithm provides an indication whether a pair of granularities could be interesting to examine together, called a harmony or clash.

 The time granularities can used to create visualizations of the data to explore for periodicities, associations and anomalies. The granularities can be considered to be categorical variables (ordered and unordered) which induces a grouping of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables. A recommendation of appropriate distribution display is provided.

 The methods are implemented in the open source R package gravitas. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow( FIXME REF), and the probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk]. 
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---
